---
title: Multi Currency Swap Pricing with SwapPricer
author: Davide Magno
date: '`r Sys.Date()`'
slug: multicurrency-swap-pricing
categories:
  - Finance
tags:
  - Coding
  - SwapPricer
  - R
  - Toolbox
keywords:
  - tech
---

```{r, message = FALSE, echo = FALSE}
# library(devtools)
devtools::install_github("DavideMagno/SwapPricer")
`%>%` <- magrittr::`%>%`
`%<>%` <- magrittr::`%<>%`
```

I have released a new version of the package **SwapPricer** on GitHub 
[here](https://github.com/DavideMagno/SwapPricer). 

As we are now at version 1.0.1 the toolbox is able to price using just a 
one-curve framework but is able to price multiple currencies (ie. CHF, EUR, GBP,
JPY and USD) and any convention in terms of coupon frequency, day count convention.

**We are working to introduce OIS Discounting in the next releases**

## SwapPricer: instructions for use

To run a multi-currency swap portfolio valuation you need the following three 
"ingredients":

1) A table with the characteristics of the swap, like the following one

    ```{r, echo = FALSE, message = FALSE, warning = FALSE}
    SwapPricer::swap.basket %>% 
      dplyr::mutate_at(dplyr::vars("notional"), ~formattable::comma(., digits = 0)) %>%
      dplyr::mutate_at(dplyr::vars("strike"), formattable::percent) %>% 
      knitr::kable(caption = "Swap portfolio", "html", align = "c") %>% 
      kableExtra::scroll_box(width = "690px")
    ```
    These are the columns that have to be in the 

    * **ID** (character) is a custom ID for the swap
    * **currency** (character) as per the ISO 4217 standard 
    * **notional** (numeric)
    * **start.date** (Date)
    * **maturity.date** (Date)
    * **strike** (numeric)
    * **type** (string) either *payer* or *receiver*. It is not case-sensitive.
    * **standard** (logical) 

    Only if the field *standard* is set to FALSE then we need four additional fields.

    An example of portfolio is provided with the package and is called *swap.basket*.

    * **time.unit.pay** and **time.unit.receive** (integer) are the number of months for the frequency of the leg (ie. monthly would have a time.unit of 1, quarterly of 3, semiannual of 6 and annual of 12)
    * **dcc.pay** and *dc..receive* (character) as per the *fmdates* helper

2) The date at which the swaps are being priced

3) As many interest rate lists as per the currencies in the swap portfolio.
The list is made of a string with the code of the currency and a with a tibble
with the discounting factor curve with two columns: Dates and Discount Factors (df). 
Here is the structure of interest rate object:

    ```{r, message = FALSE}
    str(SwapPricer::EUR.curves)
    ```
    Where the *discount* tibble of the list has the same characteristics of the `df.table` object described in [this post](https://www.curiousfrm.com/2019/04/real-world-tidy-interest-rate-swap-pricing/) 
    For your reference, I provided three curve objects with the package: *EUR.curves*, *GBP.curves* and
*USD.curves*

Now we can price our test multi-currency and non-standard swap portfolio by simply running `SwapPortfolioPricing` function with the three items in the order of description and appending as many curves as the number of currencies in the book.

A sample code is shown here:

```{r}
today <- lubridate::ymd(20190414)

results <- SwapPricer::SwapPortfolioPricing(SwapPricer::swap.basket, 
                                            today, 
                                            SwapPricer::EUR.curves, 
                                            SwapPricer::GBP.curves, 
                                            SwapPricer::USD.curves)
```

and these are the results in table format:

```{r, echo = FALSE}
results %>% 
  dplyr::mutate_at(dplyr::vars("clean.mv","dirty.mv","accrual.pay","accrual.receive","pv01"),
                   formattable::comma) %>%
  dplyr::mutate_at(dplyr::vars("par"), formattable::percent) %>%
  knitr::kable(caption = "Pricing results", "html") %>%
  kableExtra::scroll_box(width = "730px")
```

## So long RQuantLib

Apart from the new definition of the curve object, most of the changes have happened "behind the scenes". One thing that I had to do to make the deployment easier on all the OSs was to remove any reference to Quantlib. I am a huge fan of the RQuantLib package, but I realised I was using it just for adjusting dates for holidays. So I decided to take a different approach:

1. I downloaded all the holidays and saved it as internal objects in my package
2. I created a very simple routine to adjust for holidays, as in the following piece of code:

```{r}
AdvanceDate <- function(dates, currency, eom.check) {
  holiday <- holidays[[currency]]
  check <- TRUE %in% ((dates %in% holiday) |
                        (weekdays(dates) %in% "Saturday") |
                        (weekdays(dates) %in% "Sunday"))
  if (check) {
    repeat {
      if (eom.check) {
        dates <- dplyr::case_when((dates %in% holiday) ~ dates - 1,
                                      (weekdays(dates) %in% "Saturday") ~ dates - 1,
                                      (weekdays(dates) %in% "Sunday") ~ dates - 2,
                                      TRUE ~ dates)
      } else {
        dates <- dplyr::case_when((dates %in% holiday) ~ dates + 1,
                                      (weekdays(dates) %in% "Saturday") ~ dates + 2,
                                      (weekdays(dates) %in% "Sunday") ~ dates + 1,
                                      TRUE ~ dates)
      }
      check <- TRUE %in% ((dates %in% holiday) |
                            (weekdays(dates) %in% "Saturday") |
                            (weekdays(dates) %in% "Sunday"))
      if (!check) {
        return(dates)
      }
    }
  } else {
    return(dates)
  }
}
```

You can see that adjusting for holidays is a recursive exercise: let's assume we have Easter Sunday in the cashflow schedule. It automatically gets adjusted to Monday, but isince Easter Monday is also an holiday, the function will run a second time to further advance to Tuesday. 

The performance of the function is guaranteed by the fact that the **dplyr::case_when** function is vectorised. In fact, the overall procedure takes almost the same time to run as with the RQuantLib function.
